---
title: "Intro to Geospatial Data"
author: "Maya Van Nuys"
date: "1/25/2021"
output: pdf_document
---



```{r}
# Setup: 
require(tidyverse)
require(ggplot2)
require(ggmap)
require(sf)

pdb_map <- read.csv("~//geospatial//pdb_map_sample.csv")

# To plot later, we need the coordinates to be numeric

pdb_map$lat <- as.numeric(pdb_map$lat)
pdb_map$long <- as.numeric(pdb_map$long)

# Creating date and coordinate variables from the existing columns:
pdb_map <- pdb_map %>%
  mutate(date = as.Date(paste(Year, Month, Day, sep = "-"))) %>%
  mutate(coordinates = paste(lat, long, sep = ", ")) 

```

# ggplot2 

## You can create basic maps with ggplot2, treating coordinate data just as you would points on a graph. The x-axis corresponds to longitude and the y-axis to latitude. 

```{r}
# Simple map using ggplot2

ggplot(pdb_map, aes(long, lat)) +
  geom_point()
```


## So now we have the coordinates of our data plotted on an x-y axis. But we do not have a background layer to tell where these points are on a world map. To do this, we will need a base layer of the world map.  

```{r}
world_map <- map_data('world')

ggplot() +
  geom_polygon(data = world_map, aes(long, lat, group= group), fill = "lightgray") +
  geom_point(data = pdb_map, aes(x = long, y = lat))
```

## Great! Now we can see the locations of the centroids for PDB maps. Notice the concentration of maps in Southeast Asia - this makes sense given that the sample covers the years 1961-1977. Now let's try some aesthetic changes:

```{r}
# What if we want to adjust coordinate point size by a variable?

# Frequency of map centroids at certain coordinates in PDBs: 
pdb_freq <- pdb_map %>%
  group_by(coordinates) %>%
  mutate(count = sum(length(coordinates)))
  
ggplot() +
  geom_polygon(data = world_map, aes(long, lat, group= group), fill = "lightgray") +
  geom_point(data = pdb_freq, aes(x = long, y = lat, size = count), color = "blue", alpha = 0.3)

```

# tmap

## tmap is similar to ggplot2, but it is tailored for making thematic maps. Like ggplot2, the package uses flexible layering to visualize spatial data. 

## tmap functions: 

* tm_shape() --> ggplot()
* tm_borders() --> geom_polygon()
* tm_lines() --> geom_line()
* tm_dots() --> geom_point()
+ tm_bubbles()
  
  
# sf

## The sf library (Simple Features) is a relatively recent library for handling spatial data in R which makes data frame manipulation easier than its predecessor library, sp. sf allows us to work with tidyverse functions. We can use sf to transform our coordinate data into spatial data for plotting in tmap without using an outside GIS software like QGIS. 
  

```{r}
require(tmap)
require(sf)

# Now we want spatial data as an sf object to use tmap. Note that we are setting the CRS here to a fairly standard one - the EPSG projection 4326 which corresponds to the WGS84 reference ellipsoid. 

map_sf <- st_as_sf(pdb_freq, coords = c("long", "lat"), crs = 4326)

# Now let's try a simple map with tmap: 

qtm(map_sf)

# Because our data only contains spatial points and no polygons, we run into the same issue as we did with ggplot2. We need a base polygon layer for the world map:

data("World") #this data comes with the tmap package

# Now we can layer our point data over the built-in world polygon data: 

tm_shape(World) +
  tm_borders() +
  tm_shape(map_sf) +
  tm_dots()

```

# Bonus: GIFs

## While not always the most useful for data analysis, tmap makes it very easy to create GIFs with maps. For the PDB maps, we can create a GIF of the map locations by year while retaining the formatting for point-size by count. [h/t to Zach Klamann for insight on GIFs] 

## To use the tmap_animation function, you will need to install ImageMagick: https://imagemagick.org/script/download.php  

```{r}
require(magick)
require(gifski)

pdb_gif <- tm_shape(World) +
  tm_borders() +
  tm_shape(map_sf) +
  tm_bubbles(size = 'count', col = "blue", alpha = 0.3) +
  tm_facets(along = "Year", free.coords = F)

tmap_animation(pdb_gif, "pdb_maps.gif", loop = TRUE, delay = 100)


```


# PRIO-GRID

## The PRIO-GRID dataset is a grid structure spatial data set with infomration on armed conflicts, socio-economic conditions, ethnic groups, physical attributes, climatic conditions, and other variables. You can download csv or spatial data from static and time-sensitive variables. You can find the data set and more information here: https://grid.prio.org/#/ 

### In the code below, we are pulling the full shape file (spatial data file) from the PRIO website and downloading it onto our computer as a zip file. Then we unzip the file into a new folder.  


```{r}
url <- "https://grid.prio.org/extensions/priogrid_cellshp.zip"

destfile <- "~\\geospatial\\output.zip"

download.file(url, destfile)

unzip("~\\geospatial\\output.zip", overwrite = T, exdir = "~\\geospatial\\output")

```


### Next, we use sf's st_read function to read the spatial data into R. Notice the "layer" argument takes the name shared by the multiple files in the new output folder (though each has a different extension). The class of the resulting object is an sf data frame. 

### Next, we can read in a csv of several variables from the PRIO data. This is necessary in order to join country ids (gid) and the associated variables with the shape file. 

```{r}
prio <- st_read("~\\geospatial\\output",
                  layer = 'priogrid_cell',
                 stringsAsFactors = F) %>%
  mutate(gid = as.character(gid))


class(prio)

prio_vars <- read.csv("prio_vars.csv") %>%
  mutate(gid = as.character(ï..gid)) %>%
  select(-ï..gid)

prio <- left_join(prio, prio_vars, by = "gid")


```


### Now we can filter for spatial data pertaining to Vietnam only. We then use tmap to visualize the PRIO data on forest coverage with the map centroids layered over. 


```{r}
vietnam <- prio %>%
  filter(gwno == 816) 

tm_shape(vietnam) +
  tm_fill(col = "forest_ih") +
  tm_shape(map_sf) +
  tm_dots()

```

## Merging

### The sf package has great functionality for combining spatial data sets. The st_join function operates essentially like merge for non-spatial data frames. In the example below, we are matching the centroid locations from PDB maps with the information from the PRIO-GRID data set. We can then use these data to test hypotheses, such as whether the locations in PDB maps tend to be more or less barren and whether this trend has changed over time.

```{r}
pdb_int <- st_join(map_sf, prio)

which(duplicated(pdb_int$MapPDBid))
pdb_int <- pdb_int[-96:-98,]

```


